---
title: "Savitzky-Golay parameter tuning — joint optimisation with PLSR"
author: Franck Albinet
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

Savitzky-Golay (SG) smoothing and differentiation is applied to MIR spectra
before modelling to suppress baseline drift and sharpen absorption features.
It has three parameters: window size `w` (number of channels), polynomial
degree `p`, and derivative order `m`. These interact non-trivially with the
number of PLS components, so they are tuned jointly with PLSR rather than
fixed by convention.

The target property here is `log10(exCs137 / totalCs137)` — the same as the
primary modelling notebook — so the parameters found here are directly
transferable.

# Setup

```{r}
#| label: libraries

library(tidyverse)
library(tidymodels)
library(prospectr)
library(resemble)
library(here)
library(plsmod)

source(here("R", "utils.R"))
source(here("R", "load.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))
```

```{r}
#| label: configuration

# --- Data -------------------------------------------------------------------
WET_PATH     <- here("data-raw", "WetData_2013_2023.csv")
SPECTRA_PATH <- here("data-raw", "1320_Jumpei_spectra_outlier_full.rds")
BY_WET       <- "ID"
BY_SPECTRA   <- "sample_id"

# --- Target property --------------------------------------------------------
TARGET <- "exCs137_totalCs137"

# --- Filtering --------------------------------------------------------------
EXCLUDE_YEARS <- c(2021, 2022, 2023)

# --- Preprocessing ----------------------------------------------------------
RESAMPLE_BY <- 5L

# --- SG tuning grid ---------------------------------------------------------
SG_W_MIN <- 5L    # must be odd
SG_W_MAX <- 15L   # must be odd
SG_P_MAX <- 3L    # constraints enforced in grid: p < w and p >= m
SG_M_MAX <- 2L    # derivative order: 0 = smoothing, 1 = first, 2 = second

# --- Splitting --------------------------------------------------------------
SEED            <- 123L
TRAIN_PROP      <- 0.75
CALIB_PROP      <- 0.65
KS_MAX_EXPL_VAR <- 0.99

# --- PLSR -------------------------------------------------------------------
MAX_NCOMP <- 50L

set.seed(SEED)
```

# Load & Prepare

```{r}
#| label: load

raw <- load_raw_data(
  wet_path     = WET_PATH,
  spectra_path = SPECTRA_PATH,
  by_wet       = BY_WET,
  by_spectra   = BY_SPECTRA
)
```

```{r}
#| label: prepare

prepared <- raw |>
  filter(
    !year %in% EXCLUDE_YEARS,
    !is.na(.data[[TARGET]]),
    .data[[TARGET]] > 0
  )
nrow(prepared)
```

# Split

Kennard-Stone sampling on spectral PCA scores ensures the training set covers
the spectral space uniformly. This matters for SG tuning: a training set that
leaves spectral gaps could favour parameters that happen to smooth over noisy
regions rather than parameters that extract genuine chemical information.

```{r}
#| label: split

split <- ks_split(prepared, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)
```

# Recipe

The SG parameters `w`, `p`, and `m` are marked as `tune()` so the grid search
can vary them. All other preprocessing steps are fixed: log₁₀ outcome
transform, spectral resampling at `r RESAMPLE_BY` cm⁻¹, and SNV scatter
correction applied before SG differentiation.

```{r}
#| label: recipe

spc_cols  <- grep("^X\\d+", names(training(split)), value = TRUE)
meta_cols <- setdiff(names(training(split)), c(TARGET, spc_cols))

rec <- recipe(training(split)) |>
  update_role(all_of(TARGET),    new_role = "outcome")   |>
  update_role(all_of(spc_cols),  new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata")  |>
  step_log(all_outcomes())          |>
  step_resample(by = RESAMPLE_BY)   |>
  step_snv()                        |>
  step_sg(w = tune(), p = tune(), m = tune())
```

# Tune

## Workflow

PLSR is used as the model for tuning because it is fast, well-suited to
high-dimensional collinear spectra, and its optimal component count is
informative: if good SG parameters require many components, the preprocessing
is doing little useful work.

```{r}
#| label: pls-workflow

pls_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(
    pls(num_comp = tune()) |>
      set_engine("mixOmics") |>
      set_mode("regression")
  )
```

## Grid

The SG grid is constructed explicitly to respect two hard constraints:
polynomial degree must be strictly less than the window size (`p < w`), and
the derivative order must not exceed the polynomial degree (`m ≤ p`). Violating
either makes the filter mathematically ill-defined. The joint grid crosses all
valid SG combinations with every PLS component count up to `MAX_NCOMP`.

```{r}
#| label: joint-grid

sg_grid <- expand.grid(
  w = seq(SG_W_MIN, SG_W_MAX, by = 2L),
  p = seq_len(SG_P_MAX),
  m = seq(0L, SG_M_MAX)
) |>
  filter(p < w, p >= m)

joint_grid <- expand_grid(
  num_comp = seq_len(MAX_NCOMP),
  sg_grid
)

nrow(joint_grid)
```

```{r}
#| label: tune-sg

sg_tuned <- tune_grid(
  pls_wf,
  resamples = validation_set(split),
  grid      = joint_grid,
  metrics   = metric_set(rmse, rsq)
)
```

# Results

## Heatmap — best RMSE by SG parameters

Each tile shows the best validation RMSE achievable with that `(w, p)`
combination, minimised over all `num_comp` values. Facets separate the three
derivative orders: `m = 0` is pure smoothing, `m = 1` is the first derivative
(removes additive baseline), `m = 2` is the second derivative (removes linear
baseline). Darker tiles indicate worse performance.

```{r}
#| label: fig-sg-heatmap

collect_metrics(sg_tuned) |>
  filter(.metric == "rmse") |>
  group_by(w, p, m) |>
  summarise(best_rmse = min(mean), .groups = "drop") |>
  ggplot(aes(x = factor(w), y = factor(p), fill = best_rmse)) +
  geom_tile() +
  geom_text(aes(label = round(best_rmse, 3)), size = 3) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  facet_wrap(~ paste0("m = ", m)) +
  labs(
    x     = "Window size (w)",
    y     = "Polynomial degree (p)",
    fill  = "Best RMSE",
    title = "Best RMSE across num_comp — by SG parameters, faceted by derivative order"
  ) +
  theme_bw()
```

## RMSE vs number of components

The curves below show how validation RMSE evolves with the number of PLS
components for every valid SG combination, again faceted by derivative order.
A combination that reaches its minimum quickly (few components) and at a low
RMSE value is preferred — it suggests the preprocessing is extracting compact,
informative features.

```{r}
#| label: fig-ncomp-by-sg

collect_metrics(sg_tuned) |>
  filter(.metric == "rmse") |>
  mutate(sg_label = glue::glue("w={w}, p={p}")) |>
  ggplot(aes(x = num_comp, y = mean, colour = sg_label, group = sg_label)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~ paste0("m = ", m)) +
  labs(
    x      = "Number of PLS components",
    y      = "RMSE (validation)",
    colour = "SG (w, p)",
    title  = "Validation RMSE by num_comp — faceted by derivative order"
  ) +
  theme_bw()
```

## Best parameters

```{r}
#| label: best-params

best <- select_best(sg_tuned, metric = "rmse")
best

glue::glue(
  "→ Update cs137_ratio.qmd:\n",
  "  SG_W <- {best$w}L\n",
  "  SG_P <- {best$p}L"
)
```
