---
title: "log10(exCs137 / total Cs137) — MIR prediction pipeline"
author: Franck Albinet
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# Setup

```{r}
#| label: libraries

library(tidyverse)
library(tidymodels)
library(prospectr)
library(resemble)
library(here)
library(plsmod)
library(rules)
library(Cubist)
library(kernlab)
library(hexbin)
library(ranger)
library(xgboost)

source(here("R", "utils.R"))
source(here("R", "load.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))

library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1L)
```

```{r}
#| label: configuration

# --- Data -------------------------------------------------------------------
WET_PATH     <- here("data-raw", "WetData_2013_2023.csv")
SPECTRA_PATH <- here("data-raw", "1320_Jumpei_spectra_outlier_full.rds")
BY_WET       <- "ID"
BY_SPECTRA   <- "sample_id"

# --- Target property --------------------------------------------------------
TARGET <- "exCs137_totalCs137"

# --- Filtering --------------------------------------------------------------
EXCLUDE_YEARS <- c(2021, 2022, 2023)

# --- Preprocessing ----------------------------------------------------------
RESAMPLE_BY <- 5L
SG_W        <- 15L   # tuned in tune_sg.qmd
SG_P        <- 3L    # tuned in tune_sg.qmd
SG_M        <- 1L    # first derivative — confirmed in tune_sg.qmd

# --- Splitting --------------------------------------------------------------
SEED            <- 123L
TRAIN_PROP      <- 0.75
CALIB_PROP      <- 0.65
KS_MAX_EXPL_VAR <- 0.99

# --- Cross-validation (pool + fixed KS test set) -----------------------------
K_FOLDS    <- 10L
CV_REPEATS <- 1L

# --- Model ------------------------------------------------------------------
MAX_NCOMP     <- 50L
GRID_SIZE     <- 20L
RF_TREES      <- 1000L
XGB_TREES     <- 500L
XGB_GRID_SIZE <- 50L

set.seed(SEED)
```

# Load Data

```{r}
#| label: load

raw <- load_raw_data(
  wet_path     = WET_PATH,
  spectra_path = SPECTRA_PATH,
  by_wet       = BY_WET,
  by_spectra   = BY_SPECTRA
)

glimpse(raw)
```

# Prepare

```{r}
#| label: prepare

prepared <- raw |>
  filter(
    !year %in% EXCLUDE_YEARS,
    !is.na(.data[[TARGET]]),
    .data[[TARGET]] > 0
  )
nrow(prepared)
```


```{r}
prepared |>
    ggplot(aes(x = exCs137_totalCs137)) +
    geom_histogram(bins = 50) +
    labs(x = "exCs137 / totalCs137", title = "Raw scale")
```

```{r}
prepared |>
    slice_min(exCs137_totalCs137, prop = 0.01) |>
    select(year, ID, field_no, Soil_type, Soil_type_describe, exCs137_totalCs137) |>
    arrange(exCs137_totalCs137)
```
# Split

## Using Kennard-Stone

```{r}
#| label: split

split <- ks_split(prepared, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)
```

## Using Kennard-Stone pool + k-fold CV

```{r}
#| label: split-pool

pool_split <- ks_pool_split(prepared, pool_prop = TRAIN_PROP, max_expl_var = KS_MAX_EXPL_VAR, seed = SEED)
```

```{r}
#| label: cv-folds

cv_folds <- vfold_cv(training(pool_split), v = K_FOLDS, repeats = CV_REPEATS, strata = TARGET)
cv_folds
```

# Recipe

```{r}
#| label: recipe

spc_cols  <- grep("^X\\d+", names(training(split)), value = TRUE)
meta_cols <- setdiff(names(training(split)), c(TARGET, spc_cols))

rec <- recipe(training(split)) |>
  update_role(all_of(TARGET),    new_role = "outcome")   |>
  update_role(all_of(spc_cols),  new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata")  |>
  step_log(all_outcomes())                 |>
  step_resample(by = RESAMPLE_BY)          |>
  step_snv()                               |>
  step_sg(w = SG_W, p = SG_P, m = SG_M) 
  # step_rm_co2()
```

```{r}
#| label: plot-spectra-fn

plot_spectra <- function(data, title = NULL, ylim = NULL) {
  wns <- get_wns(data)
  spc <- as.matrix(data[, get_wn_cols(data)])

  matplot(
    x    = wns,
    y    = t(spc),
    xlab = expression("Wavenumber /" * cm^-1),
    ylab = "Absorbance",
    xlim = c(max(wns), min(wns)),
    ylim = ylim,
    main = title,
    type = "l", lty = 1,
    col  = rgb(0.5, 0.5, 0.5, 0.3)
  )
}
```

As a visual check:

```{r}
#| label: fig-spectra-check

rec_prepped <- prep(rec)
train_baked <- bake(rec_prepped, new_data = NULL)

par(mfrow = c(1, 2))
plot_spectra(training(split), title = "Raw")
plot_spectra(train_baked,     title = "Preprocessed")
par(mfrow = c(1, 1))
```

# Workflow

## PLSR

```{r}
#| label: pls-spec

pls_spec <- pls(num_comp = tune()) |>
  set_engine("mixOmics") |>
  set_mode("regression")

pls_spec
```

```{r}
#| label: pls-workflow

pls_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(pls_spec)
```

```{r}
#| label: pls-grid

pls_grid <- tibble(num_comp = seq_len(MAX_NCOMP))
```

```{r}
#| label: pls-tune

pls_tuned <- tune_grid(
  pls_wf,
  resamples = validation_set(split),
  grid      = pls_grid,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-pls-tuning

autoplot(pls_tuned)
```

```{r}
#| label: pls-final

pls_final <- pls_wf |>
  finalize_workflow(select_best(pls_tuned, metric = "rmse")) |>
  last_fit(split)

collect_metrics(pls_final)
```

```{r}
#| label: fig-pls-obs-pred

pls_preds <- collect_predictions(pls_final)

pls_preds |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "PLSR test set  |  RMSE = {round(collect_metrics(pls_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(pls_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )

pls_preds |>
  mutate(
    pred_orig = 10^.pred,
    obs_orig  = 10^.data[[TARGET]]
  ) |>
  metrics(truth = obs_orig, estimate = pred_orig)
```

## PLSR — k-fold CV on pool

Same recipe, model spec, and grid as above (`rec`, `pls_spec`, `pls_wf`, `pls_grid`) —
only the resampling strategy changes: `num_comp` is tuned via k-fold CV on the
KS pool instead of a single validation holdout, then the final model is fit on
the full pool and scored on the same KS test set used above (`pool_split`'s test
set is identical to `split`'s test set).

```{r}
#| label: pls-tune-cv

pls_tuned_cv <- tune_grid(
  pls_wf,
  resamples = cv_folds,
  grid      = pls_grid,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-pls-tuning-cv

autoplot(pls_tuned_cv)
```

`select_by_one_std_err()` trades a bit of raw CV performance for a simpler
(lower `num_comp`) model that's within noise of the numerical optimum picked
by `select_best()`:

```{r}
#| label: pls-select-compare-cv

pls_picks_cv <- bind_rows(
  select_best(pls_tuned_cv, metric = "rmse") |>
    mutate(rule = "select_best"),
  select_by_one_std_err(pls_tuned_cv, num_comp, metric = "rmse") |>
    mutate(rule = "select_by_one_std_err")
)

collect_metrics(pls_tuned_cv) |>
  filter(.metric %in% c("rmse", "rsq")) |>
  inner_join(pls_picks_cv |> select(rule, .config), by = ".config") |>
  select(rule, num_comp, .metric, mean, std_err)
```

```{r}
#| label: pls-final-cv

pls_final_cv <- pls_wf |>
  finalize_workflow(select_by_one_std_err(pls_tuned_cv, num_comp, metric = "rmse")) |>
  last_fit(pool_split)

collect_metrics(pls_final_cv)
```

```{r}
#| label: fig-pls-obs-pred-cv

pls_preds_cv <- collect_predictions(pls_final_cv)

pls_preds_cv |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "PLSR (CV) test set  |  RMSE = {round(collect_metrics(pls_final_cv) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(pls_final_cv) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )

pls_preds_cv |>
  mutate(
    pred_orig = 10^.pred,
    obs_orig  = 10^.data[[TARGET]]
  ) |>
  metrics(truth = obs_orig, estimate = pred_orig)
```

## Cubist

```{r}
#| label: cubist-spec

cubist_spec <- cubist_rules(
  committees = tune(),
  neighbors  = tune()
) |>
  set_engine("Cubist") |>
  set_mode("regression")

cubist_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(cubist_spec)
```

```{r}
#| label: cubist-tune

cubist_grid <- grid_regular(
  committees(range = c(1L, 30L)),
  neighbors(),
  levels = c(committees = 20, neighbors = 10)
)

cubist_tuned <- tune_grid(
  cubist_wf,
  resamples = validation_set(split),
  grid      = cubist_grid,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-cubist-tuning

autoplot(cubist_tuned)
select_best(cubist_tuned, metric = "rmse")
```

```{r}
#| label: cubist-final

cubist_final <- cubist_wf |>
  finalize_workflow(select_best(cubist_tuned, metric = "rmse")) |>
  last_fit(split)

collect_metrics(cubist_final)
```

```{r}
#| label: fig-cubist-obs-pred

collect_predictions(cubist_final) |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "Cubist test set  |  RMSE = {round(collect_metrics(cubist_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(cubist_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )

collect_predictions(cubist_final) |>
  mutate(
    pred_orig = 10^.pred,
    obs_orig  = 10^.data[[TARGET]]
  ) |>
  metrics(truth = obs_orig, estimate = pred_orig)
```

```{r}
#| label: fig-cubist-vip

extract_fit_engine(cubist_final)$usage |>
  mutate(
    importance = Conditions + Model,
    wn         = as.numeric(str_remove(Variable, "^X"))
  ) |>
  filter(!is.na(wn)) |>
  ggplot(aes(x = wn, y = importance)) +
  geom_line() +
  scale_x_reverse() +
  labs(
    x = expression("Wavenumber /" * cm^-1),
    y = "Variable usage (%)"
  )
```

```{r}
collect_predictions(cubist_final) |>
    filter(.data[[TARGET]] < -5.5) |>
    left_join(
      testing(split) |>
        mutate(.row = row_number()) |>
        select(-.data[[TARGET]]),
      by = ".row"
    ) |>
    select(.row, .pred, all_of(TARGET), year, everything())
```

## Random Forest

```{r}
#| label: rf-spec

rf_spec <- rand_forest(
  mtry  = tune(),
  min_n = tune(),
  trees = RF_TREES
) |>
  set_engine("ranger") |>
  set_mode("regression")

rf_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)
```

```{r}
#| label: rf-tune

rf_tuned <- tune_grid(
  rf_wf,
  resamples = validation_set(split),
  grid      = GRID_SIZE,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-rf-tuning

autoplot(rf_tuned)
select_best(rf_tuned, metric = "rmse")
```

```{r}
#| label: rf-final

rf_final <- rf_wf |>
  finalize_workflow(select_best(rf_tuned, metric = "rmse")) |>
  last_fit(split)

collect_metrics(rf_final)
```

```{r}
#| label: fig-rf-obs-pred

collect_predictions(rf_final) |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "RF test set  |  RMSE = {round(collect_metrics(rf_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(rf_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )

collect_predictions(rf_final) |>
  mutate(
    pred_orig = 10^.pred,
    obs_orig  = 10^.data[[TARGET]]
  ) |>
  metrics(truth = obs_orig, estimate = pred_orig)
```

## XGBoost

```{r}
#| label: xgb-spec

xgb_spec <- boost_tree(
  trees          = XGB_TREES,
  mtry           = tune(),
  min_n          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  sample_size    = tune()
) |>
  set_engine("xgboost") |>
  set_mode("regression")

xgb_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(xgb_spec)
```

```{r}
#| label: xgb-tune

xgb_grid <- grid_space_filling(
  finalize(mtry(), train_baked |> select(starts_with("X"))),
  min_n(),
  tree_depth(),
  learn_rate(),
  loss_reduction(),
  sample_size = sample_prop(),
  size = XGB_GRID_SIZE
)

xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = validation_set(split),
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-xgb-tuning

autoplot(xgb_tuned)
select_best(xgb_tuned, metric = "rmse")
```

```{r}
#| label: xgb-final

xgb_final <- xgb_wf |>
  finalize_workflow(select_best(xgb_tuned, metric = "rmse")) |>
  last_fit(split)

collect_metrics(xgb_final)
```

```{r}
#| label: fig-xgb-obs-pred

collect_predictions(xgb_final) |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "XGBoost test set  |  RMSE = {round(collect_metrics(xgb_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(xgb_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )

collect_predictions(xgb_final) |>
  mutate(
    pred_orig = 10^.pred,
    obs_orig  = 10^.data[[TARGET]]
  ) |>
  metrics(truth = obs_orig, estimate = pred_orig)
```
