---
title: "log10(exCs137 / total Cs137) — Cubist global vs. per soil type"
author: Franck Albinet
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

Does soil type matter for MIR-based prediction of the exchangeable-to-total
¹³⁷Cs ratio? This notebook addresses that question with two complementary
analyses:

1. **Global Cubist model** — trained on all soil types combined, with test-set
   metrics broken down by `Soil_type_describe` to reveal where the single model
   struggles.
2. **Per-soil Cubist models** — one independent model per soil type group
   (minimum 60 samples), each with its own Kennard-Stone split and
   hyperparameter tuning.

Comparing the variable importance plots across setups should show whether
certain spectral regions are universally predictive or carry signal only within
specific soil classes.

# Setup

```{r}
#| label: libraries

library(tidyverse)
library(tidymodels)
library(prospectr)
library(resemble)
library(here)
library(rules)
library(Cubist)

source(here("R", "utils.R"))
source(here("R", "load.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))

library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1L)
```

```{r}
#| label: configuration

WET_PATH     <- here("data-raw", "WetData_2013_2023.csv")
SPECTRA_PATH <- here("data-raw", "1320_Jumpei_spectra_outlier_full.rds")
BY_WET       <- "ID"
BY_SPECTRA   <- "sample_id"

TARGET <- "exCs137_totalCs137"

EXCLUDE_YEARS <- c(2021, 2022, 2023)

RESAMPLE_BY <- 5L
SG_W        <- 15L   # tuned in 00_tune_sg.qmd
SG_P        <- 3L    # tuned in 00_tune_sg.qmd
SG_M        <- 1L    # first derivative

SEED            <- 123L
TRAIN_PROP      <- 0.80
CALIB_PROP      <- 0.70
KS_MAX_EXPL_VAR <- 0.99
MIN_GROUP_N     <- 60L   # groups below this are skipped

set.seed(SEED)
```

# Load & Prepare

```{r}
#| label: load

raw <- load_raw_data(
  wet_path     = WET_PATH,
  spectra_path = SPECTRA_PATH,
  by_wet       = BY_WET,
  by_spectra   = BY_SPECTRA
)
```

Years 2021–2023 are excluded because measurement protocols changed; samples
with missing or non-positive target values are removed before any split so
the log transform is well-defined.

```{r}
#| label: prepare

prepared <- raw |>
  filter(
    !year %in% EXCLUDE_YEARS,
    !is.na(.data[[TARGET]]),
    .data[[TARGET]] > 0
  )
nrow(prepared)
```

# Shared Cubist specification

Cubist is chosen for its interpretability: it produces a set of rule-based
linear models whose variable usage statistics give a direct measure of spectral
feature importance. The same specification and grid are reused for every model
in this notebook to keep comparisons fair.

The tuning grid searches 20 values of `committees` (1–30) and 10 values of
`neighbors` (0–9), for 200 candidates evaluated on the validation set.

```{r}
#| label: cubist-spec

cubist_spec <- cubist_rules(
  committees = tune(),
  neighbors  = tune()
) |>
  set_engine("Cubist") |>
  set_mode("regression")

cubist_grid <- grid_regular(
  committees(range = c(1L, 30L)),
  neighbors(),
  levels = c(committees = 20, neighbors = 10)
)
```

# Global model

## Split

The dataset is split using the Kennard-Stone algorithm on PCA scores of the
SNV-preprocessed spectra. This guarantees that the training set covers the
spectral space as uniformly as possible, rather than leaving gaps that
random sampling might create. The split proportions (80 % train pool / 70 %
calibration) are chosen to leave enough samples per soil type in the
per-soil section while keeping a meaningful test set.

```{r}
#| label: global-split

global_split <- ks_split(prepared, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)
```

## Recipe

The preprocessing chain follows the pipeline established in `01_cs137_ratio.qmd`:

- **log₁₀ transform** of the outcome compresses the right-skewed ratio
  distribution and stabilises variance.
- **Spectral resampling** (every `r RESAMPLE_BY` cm⁻¹) reduces collinearity
  between adjacent channels without discarding information.
- **SNV** removes multiplicative scatter effects caused by differences in
  particle size and sample packing.
- **Savitzky-Golay first derivative** (window `r SG_W`, degree `r SG_P`)
  suppresses baseline drift and sharpens spectral features; parameters were
  tuned in `00_tune_sg.qmd`.

```{r}
#| label: global-recipe

spc_cols  <- grep("^X\\d+", names(training(global_split)), value = TRUE)
meta_cols <- setdiff(names(training(global_split)), c(TARGET, spc_cols))

global_rec <- recipe(training(global_split)) |>
  update_role(all_of(TARGET),    new_role = "outcome")   |>
  update_role(all_of(spc_cols),  new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata")  |>
  step_log(all_outcomes(), base = 10) |>
  step_resample(by = RESAMPLE_BY)    |>
  step_snv()                         |>
  step_sg(w = SG_W, p = SG_P, m = SG_M)
```

## Tune

The workflow bundles the recipe and model specification. Hyperparameters are
selected by minimising RMSE on the held-out validation set.

```{r}
#| label: global-wf

global_wf <- workflow() |>
  add_recipe(global_rec) |>
  add_model(cubist_spec)
```

```{r}
#| label: global-tune

global_tuned <- tune_grid(
  global_wf,
  resamples = validation_set(global_split),
  grid      = cubist_grid,
  metrics   = metric_set(rmse, rsq)
)
```

The tuning surface below shows how RMSE and R² vary with the number of
committees and neighbours. More committees generally improve predictions by
boosting, while neighbours control the degree of instance-based correction
applied at prediction time.

```{r}
#| label: fig-global-tuning

autoplot(global_tuned)
```

## Final fit

The best hyperparameters are locked in and the model is re-trained on the
full training set before evaluation on the held-out test set — the only time
test data are touched.

```{r}
#| label: global-final

global_final <- global_wf |>
  finalize_workflow(select_best(global_tuned, metric = "rmse")) |>
  last_fit(global_split)

collect_metrics(global_final)
```

## Observed vs predicted

```{r}
#| label: fig-global-obs-pred

collect_predictions(global_final) |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "Global Cubist  |  RMSE = {round(collect_metrics(global_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(global_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )
```

## Metrics by soil type

A single global model may hide heterogeneous behaviour across soil types. The
table below breaks down test-set performance by `Soil_type_describe`, using
the same predictions — no additional model fits are involved.

```{r}
#| label: global-metrics-by-soil

collect_predictions(global_final) |>
  left_join(
    testing(global_split) |>
      mutate(.row = row_number()) |>
      select(.row, Soil_type_describe),
    by = ".row"
  ) |>
  group_by(Soil_type_describe) |>
  metrics(truth = !!sym(TARGET), estimate = .pred) |>
  print(n = Inf)
```

## Variable importance

Cubist reports variable usage as the percentage of rules and linear model
terms in which each predictor appears. Plotting this against wavenumber gives
a spectral importance profile directly interpretable in terms of known soil
absorption bands.

```{r}
#| label: fig-global-vip

extract_fit_engine(global_final)$usage |>
  mutate(wn = as.numeric(str_remove(Variable, "^X"))) |>
  filter(!is.na(wn)) |>
  arrange(wn) |>
  mutate(
    Conditions = slider::slide_dbl(Conditions, mean, .before = 5L, .after = 5L, .complete = TRUE),
    Model      = slider::slide_dbl(Model,      mean, .before = 5L, .after = 5L, .complete = TRUE)
  ) |>
  pivot_longer(c(Conditions, Model), names_to = "role", values_to = "usage") |>
  ggplot(aes(x = wn, y = usage, colour = role)) +
  geom_line() +
  scale_x_reverse() +
  labs(
    x      = expression("Wavenumber /" * cm^-1),
    y      = "Variable usage (%)",
    colour = NULL,
    title  = "Global model — variable importance"
  )
```

# Per-soil models

If soil mineralogy and organic matter content vary substantially between
`Soil_type_describe` groups, a single global model may be suboptimal for all
of them. Here, an independent Cubist model is fitted for each group with at
least `r MIN_GROUP_N` samples. Each model gets its own Kennard-Stone split
and its own hyperparameter search — there is no information sharing between
groups.

## Group sizes

```{r}
#| label: soil-group-sizes

group_sizes <- prepared |>
  count(Soil_type_describe, name = "n") |>
  arrange(desc(n))

group_sizes

skipped <- group_sizes |> filter(n < MIN_GROUP_N)
if (nrow(skipped) > 0) {
  message(
    "Skipping ", nrow(skipped), " group(s) with n < ", MIN_GROUP_N, ": ",
    paste(skipped$Soil_type_describe, collapse = ", ")
  )
}
```

## Fit per group

`fit_soil_cubist()` encapsulates the full pipeline for a single group: split,
recipe, workflow, tuning, and final fit. Applying it via `map()` over the
`group_split()` result keeps the per-group logic in one place and returns a
uniform list structure that downstream display code can iterate over.

```{r}
#| label: soil-fit-fn

# Fit a Cubist workflow on one subset; returns a named list
fit_soil_cubist <- function(grp, label = unique(grp$Soil_type_describe)) {
  split <- ks_split(grp, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)

  spc_cols  <- grep("^X\\d+", names(training(split)), value = TRUE)
  meta_cols <- setdiff(names(training(split)), c(TARGET, spc_cols))

  rec <- recipe(training(split)) |>
    update_role(all_of(TARGET),    new_role = "outcome")   |>
    update_role(all_of(spc_cols),  new_role = "predictor") |>
    update_role(all_of(meta_cols), new_role = "metadata")  |>
    step_log(all_outcomes(), base = 10) |>
    step_resample(by = RESAMPLE_BY)    |>
    step_snv()                         |>
    step_sg(w = SG_W, p = SG_P, m = SG_M)

  wf <- workflow() |>
    add_recipe(rec) |>
    add_model(cubist_spec)

  tuned <- tune_grid(
    wf,
    resamples = validation_set(split),
    grid      = cubist_grid,
    metrics   = metric_set(rmse, rsq)
  )

  final <- wf |>
    finalize_workflow(select_best(tuned, metric = "rmse")) |>
    last_fit(split)

  list(
    label = label,
    n     = nrow(grp),
    split = split,
    tuned = tuned,
    final = final
  )
}
```

```{r}
#| label: soil-fit

soil_results <- prepared |>
  group_by(Soil_type_describe) |>
  filter(n() >= MIN_GROUP_N) |>
  group_split() |>
  map(fit_soil_cubist)
```

## Results per soil type

For each group, the observed-vs-predicted plot and the spectral importance
profile are shown side by side. Key questions: does predictive accuracy
improve relative to the global model? Do the important spectral regions shift
between soil types, suggesting that different mineral or organic components
drive Cs exchangeability depending on soil class?

```{r}
#| label: fig-soil-results

walk(soil_results, function(res) {
  mets     <- collect_metrics(res$final)
  rmse_val <- mets |> filter(.metric == "rmse") |> pull(.estimate) |> round(3)
  rsq_val  <- mets |> filter(.metric == "rsq")  |> pull(.estimate) |> round(3)

  p_obs_pred <- collect_predictions(res$final) |>
    ggplot(aes(x = .pred, y = .data[[TARGET]])) +
    geom_point(alpha = 0.4) +
    geom_abline(colour = "red", linetype = "dashed") +
    coord_obs_pred() +
    labs(
      x     = "Predicted log10(exCs137 / totalCs137)",
      y     = "Observed log10(exCs137 / totalCs137)",
      title = glue::glue(
        "{res$label}  (n = {res$n})  |  RMSE = {rmse_val},  R² = {rsq_val}"
      )
    )

  p_vip <- extract_fit_engine(res$final)$usage |>
    mutate(wn = as.numeric(str_remove(Variable, "^X"))) |>
    filter(!is.na(wn)) |>
    arrange(wn) |>
    mutate(
      Conditions = slider::slide_dbl(Conditions, mean, .before = 5L, .after = 5L, .complete = TRUE),
      Model      = slider::slide_dbl(Model,      mean, .before = 5L, .after = 5L, .complete = TRUE)
    ) |>
    pivot_longer(c(Conditions, Model), names_to = "role", values_to = "usage") |>
    ggplot(aes(x = wn, y = usage, colour = role)) +
    geom_line() +
    scale_x_reverse() +
    labs(
      x      = expression("Wavenumber /" * cm^-1),
      y      = "Variable usage (%)",
      colour = NULL,
      title  = glue::glue("{res$label} — variable importance")
    )

  print(p_obs_pred)
  print(p_vip)
})
```

## Metrics summary

The table below collects test-set RMSE and R² for every per-soil model,
sorted by R². Comparing these figures against the corresponding rows in the
global-model breakdown above quantifies the gain — or lack thereof — from
soil-type stratification.

```{r}
#| label: soil-metrics-summary

map(soil_results, function(res) {
  collect_metrics(res$final) |>
    mutate(soil_type = res$label, n = res$n)
}) |>
  list_rbind() |>
  select(soil_type, n, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(desc(rsq))
```

# Lowland soil — by field type

Within Lowland soil, two distinct agricultural contexts coexist: paddy fields
(flooded rice cultivation, n = 684) and upland fields (n = 303). Flooding
cycles alter soil redox chemistry and organic matter dynamics, which could
affect both Cs exchangeability and its spectral expression. A separate Cubist
model for each field type tests whether this distinction matters for
prediction and whether the important spectral features shift between contexts.

## Group sizes

```{r}
#| label: lowland-group-sizes

prepared |>
  filter(Soil_type_describe == "Lowland soil") |>
  count(fieldtype, sort = TRUE)
```

## Fit per field type

```{r}
#| label: lowland-fit

lowland_results <- prepared |>
  filter(Soil_type_describe == "Lowland soil") |>
  group_by(fieldtype) |>
  filter(n() >= MIN_GROUP_N) |>
  group_split() |>
  map(\(grp) fit_soil_cubist(grp, label = unique(grp$fieldtype)))
```

## Results per field type

```{r}
#| label: fig-lowland-results

walk(lowland_results, function(res) {
  mets     <- collect_metrics(res$final)
  rmse_val <- mets |> filter(.metric == "rmse") |> pull(.estimate) |> round(3)
  rsq_val  <- mets |> filter(.metric == "rsq")  |> pull(.estimate) |> round(3)

  p_obs_pred <- collect_predictions(res$final) |>
    ggplot(aes(x = .pred, y = .data[[TARGET]])) +
    geom_point(alpha = 0.4) +
    geom_abline(colour = "red", linetype = "dashed") +
    coord_obs_pred() +
    labs(
      x     = "Predicted log10(exCs137 / totalCs137)",
      y     = "Observed log10(exCs137 / totalCs137)",
      title = glue::glue(
        "Lowland — {res$label}  (n = {res$n})  |  RMSE = {rmse_val},  R² = {rsq_val}"
      )
    )

  p_vip <- extract_fit_engine(res$final)$usage |>
    mutate(wn = as.numeric(str_remove(Variable, "^X"))) |>
    filter(!is.na(wn)) |>
    arrange(wn) |>
    mutate(
      Conditions = slider::slide_dbl(Conditions, mean, .before = 5L, .after = 5L, .complete = TRUE),
      Model      = slider::slide_dbl(Model,      mean, .before = 5L, .after = 5L, .complete = TRUE)
    ) |>
    pivot_longer(c(Conditions, Model), names_to = "role", values_to = "usage") |>
    ggplot(aes(x = wn, y = usage, colour = role)) +
    geom_line() +
    scale_x_reverse() +
    labs(
      x      = expression("Wavenumber /" * cm^-1),
      y      = "Variable usage (%)",
      colour = NULL,
      title  = glue::glue("Lowland — {res$label} — variable importance")
    )

  print(p_obs_pred)
  print(p_vip)
})
```

## Metrics summary

```{r}
#| label: lowland-metrics-summary

map(lowland_results, function(res) {
  collect_metrics(res$final) |>
    mutate(field_type = res$label, n = res$n)
}) |>
  list_rbind() |>
  select(field_type, n, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(desc(rsq))
```
