---
title: "log10(exCs137 / total Cs137) — PLSR global vs. per soil type + SHAP"
author: Franck Albinet
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

Does soil type matter for MIR-based prediction of the exchangeable-to-total
¹³⁷Cs ratio, and which spectral regions drive the predictions? This notebook
mirrors `03_cs137_ratio_cubist_by_soil.qmd` but replaces Cubist with PLSR and
replaces the Cubist usage-count importance (which can exceed 100 % and ignores
coefficient magnitudes) with SHAP values.

SHAP values are computed on the **post-recipe (SG-derivative) feature space**:
the training and test sets are baked through the prepped recipe before
KernelSHAP is called on the extracted parsnip model. This keeps the feature
space identical to the Cubist notebook so spectral importance profiles are
directly comparable across models.

1. **Global PLSR model** — trained on all soil types combined, with test-set
   metrics broken down by `Soil_type_describe`.
2. **Per-soil PLSR models** — one independent model per soil type group
   (minimum 60 samples), each with its own Kennard-Stone split and
   component-number tuning.

# Setup

```{r}
#| label: libraries

library(tidyverse)
library(tidymodels)
library(plsmod)
library(prospectr)
library(resemble)
library(here)
library(kernelshap)
library(shapviz)
library(future)

source(here("R", "utils.R"))
source(here("R", "load.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))

library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1L)
```

```{r}
#| label: configuration

WET_PATH     <- here("data-raw", "WetData_2013_2023.csv")
SPECTRA_PATH <- here("data-raw", "1320_Jumpei_spectra_outlier_full.rds")
BY_WET       <- "ID"
BY_SPECTRA   <- "sample_id"

TARGET <- "exCs137_totalCs137"

EXCLUDE_YEARS <- c(2021, 2022, 2023)

RESAMPLE_BY <- 5L
SG_W        <- 15L
SG_P        <- 3L
SG_M        <- 1L

SEED        <- 123L
TRAIN_PROP  <- 0.80
CALIB_PROP  <- 0.70
MIN_GROUP_N <- 60L

N_COMP_MAX <- 30L   # maximum PLS components to search
BG_N       <- 75L   # KernelSHAP background sample size
SHAP_M     <- 100L  # on-off vectors per iteration (default = 2 * n_features ≈ 1340)
SHAP_TOL   <- 0.01  # convergence tolerance (default 0.005)

set.seed(SEED)
```

# Load & Prepare

```{r}
#| label: load

raw <- load_raw_data(
  wet_path     = WET_PATH,
  spectra_path = SPECTRA_PATH,
  by_wet       = BY_WET,
  by_spectra   = BY_SPECTRA
)
```

```{r}
#| label: prepare

prepared <- raw |>
  filter(
    !year %in% EXCLUDE_YEARS,
    !is.na(.data[[TARGET]]),
    .data[[TARGET]] > 0
  )
nrow(prepared)
```

# Shared PLSR specification

PLSR handles the high collinearity of MIR spectra natively by projecting
predictors onto latent components that maximise covariance with the response.
The number of components is the only hyperparameter; the grid searches even
values from 2 to `r N_COMP_MAX`.

```{r}
#| label: pls-spec
pls_spec <- pls(num_comp = tune()) |>
  set_engine("mixOmics") |>
  set_mode("regression")

pls_grid <- tibble(num_comp = seq(2L, N_COMP_MAX, by = 2L))
```

# Shared helpers

`compute_shap()` bakes a data split through the prepped recipe, draws a
background from the training set, and returns a `shapviz` object for the test
set. `plot_shap_spectrum()` converts that object into a spectral importance
profile (mean |SHAP| vs wavenumber) comparable to the Cubist usage plots.

```{r}
#| label: helpers

# Bake split; run KernelSHAP on test set; returns shapviz object
compute_shap <- function(last_fit_obj, split) {
  rec_p   <- extract_workflow(last_fit_obj) |> extract_recipe(estimated = TRUE)
  model   <- extract_fit_parsnip(last_fit_obj)

  train_b <- bake(rec_p, new_data = training(split),  all_predictors())
  test_b  <- bake(rec_p, new_data = testing(split),   all_predictors())

  bg <- slice_sample(train_b, n = BG_N)
  pf <- function(m, x, ...) predict(m, new_data = x)$.pred

  # multisession is safe here: workers only call predict() on baked data,
  # never see custom recipe steps
  plan(multisession, workers = parallel::detectCores() - 1L)
  on.exit(plan(sequential))

  ks <- kernelshap(
    model,
    X        = test_b,
    bg_X     = bg,
    pred_fun = pf,
    m        = SHAP_M,
    tol      = SHAP_TOL,
    parallel = TRUE
  )
  shapviz(ks)
}

# Plot mean |SHAP| as a line against wavenumber
plot_shap_spectrum <- function(sv, title) {
  as.data.frame(sv$S) |>
    summarise(across(everything(), \(x) mean(abs(x)))) |>
    pivot_longer(everything(), names_to = "variable", values_to = "mean_abs_shap") |>
    mutate(wn = as.numeric(str_remove(variable, "^X"))) |>
    filter(!is.na(wn)) |>
    ggplot(aes(x = wn, y = mean_abs_shap)) +
    geom_line() +
    scale_x_reverse() +
    labs(
      x     = expression("Wavenumber /" * cm^-1),
      y     = "Mean |SHAP|",
      title = title
    )
}
```

# Global model

## Split

```{r}
#| label: global-split

global_split <- ks_split(prepared, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)
```

## Recipe

```{r}
#| label: global-recipe

spc_cols  <- grep("^X\\d+", names(training(global_split)), value = TRUE)
meta_cols <- setdiff(names(training(global_split)), c(TARGET, spc_cols))

global_rec <- recipe(training(global_split)) |>
  update_role(all_of(TARGET),    new_role = "outcome")   |>
  update_role(all_of(spc_cols),  new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata")  |>
  step_log(all_outcomes(), base = 10) |>
  step_resample(by = RESAMPLE_BY)    |>
  step_snv()                         |>
  step_sg(w = SG_W, p = SG_P, m = SG_M)
```

## Tune

```{r}
#| label: global-wf

global_wf <- workflow() |>
  add_recipe(global_rec) |>
  add_model(pls_spec)
```

```{r}
#| label: global-tune

global_tuned <- tune_grid(
  global_wf,
  resamples = validation_set(global_split),
  grid      = pls_grid,
  metrics   = metric_set(rmse, rsq)
)
```

```{r}
#| label: fig-global-tuning

autoplot(global_tuned)
```

## Final fit

```{r}
#| label: global-final

global_final <- global_wf |>
  finalize_workflow(select_best(global_tuned, metric = "rmse")) |>
  last_fit(global_split)

collect_metrics(global_final)
```

## Observed vs predicted

```{r}
#| label: fig-global-obs-pred

collect_predictions(global_final) |>
  ggplot(aes(x = .pred, y = .data[[TARGET]])) +
  geom_point(alpha = 0.4) +
  geom_abline(colour = "red", linetype = "dashed") +
  coord_obs_pred() +
  labs(
    x     = "Predicted log10(exCs137 / totalCs137)",
    y     = "Observed log10(exCs137 / totalCs137)",
    title = glue::glue(
      "Global PLSR  |  RMSE = {round(collect_metrics(global_final) |> filter(.metric == 'rmse') |> pull(.estimate), 3)},",
      "  R² = {round(collect_metrics(global_final) |> filter(.metric == 'rsq') |> pull(.estimate), 3)}"
    )
  )
```

## Metrics by soil type

```{r}
#| label: global-metrics-by-soil

collect_predictions(global_final) |>
  left_join(
    testing(global_split) |>
      mutate(.row = row_number()) |>
      select(.row, Soil_type_describe),
    by = ".row"
  ) |>
  group_by(Soil_type_describe) |>
  metrics(truth = !!sym(TARGET), estimate = .pred) |>
  print(n = Inf)
```

## SHAP importance

KernelSHAP is run on the `r BG_N`-sample training background and the full
test set, both baked through the prepped recipe. Mean absolute SHAP by
wavenumber is the spectral importance profile: unlike Cubist usage counts, the
values are bounded, magnitude-weighted, and can be compared directly across
models.

```{r}
#| label: global-shap

global_sv <- compute_shap(global_final, global_split)
```

```{r}
#| label: fig-global-shap

plot_shap_spectrum(global_sv, "Global PLSR — mean |SHAP| by wavenumber")
```

# Per-soil models

## Group sizes

```{r}
#| label: soil-group-sizes

group_sizes <- prepared |>
  count(Soil_type_describe, name = "n") |>
  arrange(desc(n))

group_sizes

skipped <- group_sizes |> filter(n < MIN_GROUP_N)
if (nrow(skipped) > 0) {
  message(
    "Skipping ", nrow(skipped), " group(s) with n < ", MIN_GROUP_N, ": ",
    paste(skipped$Soil_type_describe, collapse = ", ")
  )
}
```

## Fit per group

```{r}
#| label: soil-fit-fn

# Fit a PLSR workflow on one subset; returns a named list
fit_soil_pls <- function(grp, label = unique(grp$Soil_type_describe)) {
  split <- ks_split(grp, train_prop = TRAIN_PROP, calib_prop = CALIB_PROP)

  spc_c  <- grep("^X\\d+", names(training(split)), value = TRUE)
  meta_c <- setdiff(names(training(split)), c(TARGET, spc_c))

  rec <- recipe(training(split)) |>
    update_role(all_of(TARGET),  new_role = "outcome")   |>
    update_role(all_of(spc_c),   new_role = "predictor") |>
    update_role(all_of(meta_c),  new_role = "metadata")  |>
    step_log(all_outcomes(), base = 10) |>
    step_resample(by = RESAMPLE_BY)    |>
    step_snv()                         |>
    step_sg(w = SG_W, p = SG_P, m = SG_M)

  wf <- workflow() |>
    add_recipe(rec) |>
    add_model(pls_spec)

  tuned <- tune_grid(
    wf,
    resamples = validation_set(split),
    grid      = pls_grid,
    metrics   = metric_set(rmse, rsq)
  )

  final <- wf |>
    finalize_workflow(select_best(tuned, metric = "rmse")) |>
    last_fit(split)

  list(label = label, n = nrow(grp), split = split, tuned = tuned, final = final)
}
```

```{r}
#| label: soil-fit

soil_results <- prepared |>
  group_by(Soil_type_describe) |>
  filter(n() >= MIN_GROUP_N) |>
  group_split() |>
  map(fit_soil_pls)
```

## SHAP per soil type

```{r}
#| label: soil-shap

soil_sv <- map(soil_results, \(res) compute_shap(res$final, res$split))
```

## Results per soil type

```{r}
#| label: fig-soil-results

walk2(soil_results, soil_sv, function(res, sv) {
  mets     <- collect_metrics(res$final)
  rmse_val <- mets |> filter(.metric == "rmse") |> pull(.estimate) |> round(3)
  rsq_val  <- mets |> filter(.metric == "rsq")  |> pull(.estimate) |> round(3)

  p_obs_pred <- collect_predictions(res$final) |>
    ggplot(aes(x = .pred, y = .data[[TARGET]])) +
    geom_point(alpha = 0.4) +
    geom_abline(colour = "red", linetype = "dashed") +
    coord_obs_pred() +
    labs(
      x     = "Predicted log10(exCs137 / totalCs137)",
      y     = "Observed log10(exCs137 / totalCs137)",
      title = glue::glue(
        "{res$label}  (n = {res$n})  |  RMSE = {rmse_val},  R² = {rsq_val}"
      )
    )

  p_shap <- plot_shap_spectrum(sv, glue::glue("{res$label} — mean |SHAP|"))

  print(p_obs_pred)
  print(p_shap)
})
```

## Metrics summary

```{r}
#| label: soil-metrics-summary

map(soil_results, function(res) {
  collect_metrics(res$final) |>
    mutate(soil_type = res$label, n = res$n)
}) |>
  list_rbind() |>
  select(soil_type, n, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(desc(rsq))
```

# Lowland soil — by field type

## Group sizes

```{r}
#| label: lowland-group-sizes

prepared |>
  filter(Soil_type_describe == "Lowland soil") |>
  count(fieldtype, sort = TRUE)
```

## Fit per field type

```{r}
#| label: lowland-fit

lowland_results <- prepared |>
  filter(Soil_type_describe == "Lowland soil") |>
  group_by(fieldtype) |>
  filter(n() >= MIN_GROUP_N) |>
  group_split() |>
  map(\(grp) fit_soil_pls(grp, label = unique(grp$fieldtype)))
```

## SHAP per field type

```{r}
#| label: lowland-shap

lowland_sv <- map(lowland_results, \(res) compute_shap(res$final, res$split))
```

## Results per field type

```{r}
#| label: fig-lowland-results

walk2(lowland_results, lowland_sv, function(res, sv) {
  mets     <- collect_metrics(res$final)
  rmse_val <- mets |> filter(.metric == "rmse") |> pull(.estimate) |> round(3)
  rsq_val  <- mets |> filter(.metric == "rsq")  |> pull(.estimate) |> round(3)

  p_obs_pred <- collect_predictions(res$final) |>
    ggplot(aes(x = .pred, y = .data[[TARGET]])) +
    geom_point(alpha = 0.4) +
    geom_abline(colour = "red", linetype = "dashed") +
    coord_obs_pred() +
    labs(
      x     = "Predicted log10(exCs137 / totalCs137)",
      y     = "Observed log10(exCs137 / totalCs137)",
      title = glue::glue(
        "Lowland — {res$label}  (n = {res$n})  |  RMSE = {rmse_val},  R² = {rsq_val}"
      )
    )

  p_shap <- plot_shap_spectrum(sv, glue::glue("Lowland — {res$label} — mean |SHAP|"))

  print(p_obs_pred)
  print(p_shap)
})
```

## Metrics summary

```{r}
#| label: lowland-metrics-summary

map(lowland_results, function(res) {
  collect_metrics(res$final) |>
    mutate(field_type = res$label, n = res$n)
}) |>
  list_rbind() |>
  select(field_type, n, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(desc(rsq))
```
